Abstractions for Multi-Sorted Substitutions

### 2024-03-07 paper/paper.tex
Refactored paper for ITP 2024

% \documentclass[sigplan, screen]{acmart}
% \documentclass[sigplan, screen, anonymous, review, authordraft]{acmart}
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\documentclass[sigplan,10pt, anonymous]{acmart}

@@

\bibliographystyle{ACM-Reference-Format}

\title{Abstractions for Multi-Sorted Substitutions}

\author{Hannes Saffrich}
\orcid{0000-0002-1825-0097}                 %% \orcid is optional
\affiliation{
  % \department{Department1}                %% \department is recommended
  \institution{University of Freiburg}      %% \institution is required
  \country{Germany}                         %% \country is recommended
}
\email{saffrich@informatik.uni-freiburg.de} %% \email is recommended

@@

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material.
  The full framework is available online.\footnote{See supplementary material. In the final version here will be a link to github.}

@@

  For example, the substitution, which replaces the term variable 0 with the
  identity function and the type variable 1 with the bottom type, is defined
  as followed:
  \ExampleCode\FExampleSub

@@

  \subsection{The Need for Renamings}
  \label{sec:maps:renamings}
  A problem with mechanizing languages with DeBruijn-indices is that
  the operation of applying a substitution \ACode{σ} to a term
  \ACode{t} cannot be defined directly by structural recursion on
  \ACode{t}.
  If \ACode{t} is a term like \ACode{\ACon{λx} e},
  where the subterm \ACode{e} is under a binder, then the variables
  in \ACode{e} are shifted, i.e.\ the variable \ACode{\ACon{zero}} refers
  to the variable from the binder, whereas a variable \ACode{\ACon{suc} x}
  refers to the variable \ACode{x} from outside of the binder.
  As the terms contained in \ACode{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \ACode{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \ACode{x} to the term \ACode{\ACon{`} (\ACon{suc} x)}.
  However, applying this shifting substitution to the terms in
  \ACode{σ} is not structurally recursive, as the terms in \ACode{σ}
  are not subterms of \ACode{\ACon{λx} e}.

  The usual workaround is to first define renamings,
  i.e.\ substitutions that replace variables with variables.
  In contrast to general substitutions, renamings can be easily
  shifted: they only contain variables \ACode{x}, which can be
  shifted by simply taking their successor \ACode{\ACon{suc} x}.
  As the shifting substitution is a renaming, we can then define substitution,
  where we apply the shifting \emph{renaming} to the terms in \ACode{σ} when
  the substitution needs to go under a binder.

  % The usual workaround is to define renamings first and use them
  % to weaken the terms in substitutions when going under binders.
  While this keeps lemmas simple (requiring only structural induction),
  it duplicates the amount of work as all operations and lemmas have
  to be defined and proved for both renamings and substitutions.

@@

  Building on the operations and axioms of a \ACode{\ACon{Kit}}, we
  derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the \ACode{\ACon{Kit}} provides the
    necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

@@

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term\\ \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

@@

  \section{Typing}
  \label{sec:typing}
  Now that we have a notion of types and type context, we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping
  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of the lambda abstraction, which
    weakens the type of the codomain \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-typed due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

@@

  \section{Subject Reduction}
  \label{sec:subjectreduction}

  To finish our running example, we define a standard small-step
  operational semantics for System F with full reduction and prove subject reduction:
  \AppCode\FReduction
  \vspace{0mm}
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  We only list the cases for $\beta$-reduction, as the other cases
  follow directly from the induction hypothesis:
  \begin{itemize}
  \item The $\beta$-reduction rule for type-application, \ACode{\ACon{β-Λ}},
    is a direct consequence of substitution-preserves-typing.
  \item The $\beta$-reduction rule for value-application is almost the same,
    but requires us to additionally show that the type is not affected by the
    expression substitution.
    % This is a direct consequence of the weakening used in \ACode{\ACon{⊢λ}}.
  \end{itemize}

>>>>

% PAGE LIMIT: 16

% \documentclass[sigplan, screen]{acmart}
% \documentclass[sigplan, screen, anonymous, review, authordraft]{acmart}
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[sigplan,10pt, anonymous]{acmart}
\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}

@@

% \bibliographystyle{ACM-Reference-Format}
\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Abstractions for Multi-Sorted Substitutions}
\author{Hannes Saffrich}{University of Freiburg, Germany}{saffrich@informatik.uni-freiburg.de}{https://orcid.org/0009-0004-7014-754X}{}
\authorrunning{H. Saffrich}
\Copyright{CC-BY}
\keywords{Agda, Metatheory, Framework}
% \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
% \acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \author{Hannes Saffrich}
% \orcid{0000-0002-1825-0097}                 %% \orcid is optional
% \affiliation{
%   % \department{Department1}                %% \department is recommended
%   \institution{University of Freiburg}      %% \institution is required
%   \country{Germany}                         %% \country is recommended
% }
% \email{saffrich@informatik.uni-freiburg.de} %% \email is recommended

@@

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011039.10011040</concept_id>
       <concept_desc>Software and its engineering~Syntax</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10002990</concept_id>
       <concept_desc>Theory of computation~Logic and verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Software and its engineering~Syntax}
\ccsdesc[500]{Theory of computation~Logic and verification}

@@

  In this paper we present a simplified version of the actual
  framework, focusing on the core concepts. We present all necessary
  definitions and lemmas, but omit some proofs for the sake of
  space. The omitted proofs can be found in Appendix~\ref{sec:proofs}
  and in the supplementary material.
  The full framework is available online.\footnote{See supplementary material. In the final version, this text will be replaced by a link to github.}

@@

  % For example, the substitution, which replaces the term variable 0 with the
  % identity function and the type variable 1 with the bottom type, is defined
  % as followed:
  % \ExampleCode\FExampleSub

@@

  % \subsection{The Need for Renamings}
  % \label{sec:maps:renamings}
  % A problem with mechanizing languages with DeBruijn-indices is that
  % the operation of applying a substitution \ACode{σ} to a term
  % \ACode{t} cannot be defined directly by structural recursion on
  % \ACode{t}.
  % If \ACode{t} is a term like \ACode{\ACon{λx} e},
  % where the subterm \ACode{e} is under a binder, then the variables
  % in \ACode{e} are shifted, i.e.\ the variable \ACode{\ACon{zero}} refers
  % to the variable from the binder, whereas a variable \ACode{\ACon{suc} x}
  % refers to the variable \ACode{x} from outside of the binder.
  % As the terms contained in \ACode{σ} may themselves contain free variables,
  % they need to be shifted, too, before they can be substituted
  % for variables in \ACode{e}. 
  % Shifting the variables in a term is itself a substitution, which maps
  % each variable \ACode{x} to the term \ACode{\ACon{`} (\ACon{suc} x)}.
  % However, applying this shifting substitution to the terms in
  % \ACode{σ} is not structurally recursive, as the terms in \ACode{σ}
  % are not subterms of \ACode{\ACon{λx} e}.

  % The usual workaround is to first define renamings,
  % i.e.\ substitutions that replace variables with variables.
  % In contrast to general substitutions, renamings can be easily
  % shifted: they only contain variables \ACode{x}, which can be
  % shifted by simply taking their successor \ACode{\ACon{suc} x}.
  % As the shifting substitution is a renaming, we can then define substitution,
  % where we apply the shifting \emph{renaming} to the terms in \ACode{σ} when
  % the substitution needs to go under a binder.

  % % The usual workaround is to define renamings first and use them
  % % to weaken the terms in substitutions when going under binders.
  % While this keeps lemmas simple (requiring only structural induction),
  % it duplicates the amount of work as all operations and lemmas have
  % to be defined and proved for both renamings and substitutions.

@@

  Building on the operations and axioms of a \ACode{\ACon{Kit}}, we
  define the usual operations for renamings and substitutions:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the \ACode{\ACon{Kit}} provides the
    necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

@@

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

@@

  \section{Typing}
  \label{sec:typing}
  Now that we have a notion of types and type context, we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping

  % \begin{minipage}[b]{0.5\linewidth}
  % \AppCode\FTypingI
  % \end{minipage}
  % \begin{minipage}[b]{0.5\linewidth}
  % \AppCode\FTypingII
  % \end{minipage}

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of the lambda abstraction, which
    weakens the type of the codomain \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-typed due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

@@

  % \section{Subject Reduction}
  % \label{sec:subjectreduction}

  % To finish our running example, we define a standard small-step
  % operational semantics for System F with full reduction and prove subject reduction:
  % \AppCode\FReduction
  % \vspace{0mm}
  % \AppCode\FSubjectReduction
  % \AppCode\FSubjectReductionProofInteresting
  % We only list the cases for $\beta$-reduction, as the other cases
  % follow directly from the induction hypothesis:
  % \begin{itemize}
  % \item The $\beta$-reduction rule for type-application, \ACode{\ACon{β-Λ}},
  %   is a direct consequence of substitution-preserves-typing.
  % \item The $\beta$-reduction rule for value-application is almost the same,
  %   but requires us to additionally show that the type is not affected by the
  %   expression substitution.
  %   % This is a direct consequence of the weakening used in \ACode{\ACon{⊢λ}}.
  % \end{itemize}
